---
title: "Export a data set in SCE format"
author: "Roman Hillje"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Export a data set in SCE format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

Here, I explain how one can export a data set stored in the [`SingleCellExperiment`/`SCE`](https://bioconductor.org/packages/release/bioc/html/SingleCellExperiment.html) format to visualize it in Cerebro.

The workflow used here is loosely put together from the [`scran`](https://bioconductor.org/packages/release/bioc/vignettes/scran/inst/doc/scran.html) and [`scater`](https://bioconductor.org/packages/3.11/bioc/vignettes/scater/inst/doc/overview.html) examples.
Do not use this workflow as a reference for how to properly process your data.

# Installation

Before starting the workflow, we need to install `cerebroApp`, as well as the [`scran`](https://bioconductor.org/packages/release/bioc/html/scran.html), [`scater`](https://bioconductor.org/packages/3.11/bioc/html/scater.html) and [`SingleR`](https://www.bioconductor.org/packages/release/bioc/html/SingleR.html) packages, which are not installed as dependencies of cerebroApp because they are only necessary if you want/need to pre-process your scRNA-seq data.
If you just want to launch the Cerebro user interface, e.g. because you already have the pre-processed data, you don't need those packages.

```{r installation, message=FALSE, warning=FALSE}
if ( 'BiocManager' %in% installed.packages() == FALSE ) install.packages('BiocManager')

library(BiocManager)

if ( 'cerebroApp' %in% installed.packages() == FALSE ) install('cerebroApp')
if ( 'scran' %in% installed.packages() == FALSE ) install('scran')
if ( 'scater' %in% installed.packages() == FALSE ) install('scater')
if ( 'SingleR' %in% installed.packages() == FALSE ) install('SingleR')
if ( 'monocle' %in% installed.packages() == FALSE ) install('monocle')
```

# Load packages

Apart from the packages we just installed, we will also load the `dplyr` package and set a seed to make our analysis reproducible.

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(scran)
library(scater)
library(SingleR)
library(cerebroApp)

options(width = 100)
set.seed(1234567)
```

# Load count data

In this example workflow, we will load a small transcript count table from the [`Seurat`](https://satijalab.org/seurat/) package containing 80 cells and 230 genes.
If you want to try the workflow with a real data set, you can find some transcript count matrices on the [Cerebro GitHub repo](https://github.com/romanhaa/Cerebro/tree/master/examples).

```{r load_data}
pbmc_counts <- read.table(
    file = system.file('extdata', 'pbmc_raw.txt', package = 'Seurat'),
    as.is = TRUE
  ) %>% as.matrix()
```

# Pre-processing with `scran` and `scater`

Now, we create a `SingleCellExperiment` object and filter out cells with less than `50` transcripts or fewer than `10` expressed genes.
Those cut-offs are only reasonable for this example data set and will likely need to be adjusted in a real data set.
Then, we...

* remove genes expressed in less than 5 cells,
* log-normalize transcript counts,
* identify highly variably genes,
* perform principal component analysis (PCA).

```{r pre_processing, warning=FALSE}
sce <- SingleCellExperiment::SingleCellExperiment(
  assays = list(counts = pbmc_counts)
)

qc_metrics <- perCellQCMetrics(sce)
colData(sce) <- cbind(colData(sce), qc_metrics)

cells_to_keep <- sce$sum >= 50 & sce$detected >= 10
sce <- sce[,cells_to_keep]

genes_to_keep <- nexprs(sce, byrow=TRUE) >= 5
sce <- sce[genes_to_keep,]

sce <- logNormCounts(sce)

variable_genes <- modelGeneVar(sce)
highly_variable_genes <- getTopHVGs(variable_genes, n = 50)

sce <- runPCA(sce)
```

# Clustering

Then, we identify clusters using a graph-based approach.

```{r clustering}
graph <- buildSNNGraph(sce, use.dimred = "PCA")
cluster <- igraph::cluster_walktrap(graph)$membership
sce$cluster <- factor(cluster)
```

# Dimensional reduction

Next, we use the UMAP technique to generate two- and three-dimensional projections.

```{r dimensional_reduction, warning=FALSE}
sce <- runUMAP(
  sce,
  name = 'UMAP',
  ncomponents = 2,
)

sce <- runUMAP(
  sce,
  name = 'UMAP_3D',
  ncomponents = 3,
)
```

# Curate meta data

This example data set consists of a single sample so we just add that name to the meta data.
Moreover, in order to be able to understand how we did the analysis later, we add some meta data to the `misc` slot of our `SCE` object, e.g. an experiment name and the species.
The info we store there will later be collected for visualization in Cerebro.

```{r prepare_meta_data}
sce$sample <- factor('pbmc', levels = 'pbmc')

sce@metadata$experiment <- list(
  experiment_name = 'pbmc',
  organism = 'hg',
  date_of_analysis = Sys.Date()
)

sce@metadata$parameters <- list(
  gene_nomenclature = 'gene_name',
  discard_genes_expressed_in_fewer_cells_than = 10,
  keep_mitochondrial_genes = TRUE,
  variables_to_regress_out = 'nUMI'
)

sce@metadata$parameters$filtering <- list(
  UMI_min = 50,
  UMI_max = Inf,
  genes_min = 10,
  genes_max = Inf
)

sce@metadata$technical_info$cerebroApp_version <- utils::packageVersion('cerebroApp')
sce@metadata$technical_info$scran <- utils::packageVersion('scran')
sce@metadata$technical_info$scater <- utils::packageVersion('scater')
sce@metadata$technical_info <- list(
  'R' = capture.output(devtools::session_info())
)
```

# Assign cell types

Using the [`SingleR`](https://www.bioconductor.org/packages/release/bioc/html/SingleR.html) package, we can get a suggestion of cell type for each cell.
Apart from the suggested cell type, we also extract the score that might help us to understand how confident the assignment was.

```{r cell_type_assignment}
singler_ref <- BlueprintEncodeData()

singler_results_blueprintencode_main <- SingleR(
  test = logcounts(sce),
  ref = singler_ref,
  labels = singler_ref@colData@listData$label.main
)

sce$cell_type_singler_blueprintencode_main <- singler_results_blueprintencode_main@listData$labels

singler_scores <- singler_results_blueprintencode_main@listData$scores %>%
  as_tibble() %>%
  dplyr::mutate(assigned_score = NA)

for ( i in seq_len(nrow(singler_scores)) ) {
  singler_scores$assigned_score[i] <- singler_scores[[singler_results_blueprintencode_main@listData$labels[i]]][i]
}

sce$cell_type_singler_blueprintencode_main_score <- singler_scores$assigned_score
```

##Identify marker genes

Marker genes are genes which are particularly strongly or weakly expressed in a given cell population, e.g. a cluster.
This information can help to distinguish the role of different cell groups in the data set.
Here, marker genes are identified for clusters and cell types using the `findMarkers()` function from the `scran` package.
The results are separate tables for the respective subgroups, which we merge together in order to make them compatible with Cerebro.

## Clusters

```{r marker_genes_clusters}
sce@metadata$marker_genes <- list()
sce@metadata$marker_genes$scran <- list()

marker_genes_cluster <- findMarkers(sce, sce$cluster)

for ( i in seq_along(marker_genes_cluster) ) {
  marker_genes_cluster[[i]] <- as.data.frame(marker_genes_cluster[[i]]) %>%
    mutate(
      cluster = names(marker_genes_cluster)[i],
      gene = rownames(.)
    )
  marker_genes_cluster[[i]][[paste0('logFC.', names(marker_genes_cluster)[i])]] <- 0
}

marker_genes_cluster <- do.call(bind_rows, marker_genes_cluster) %>%
  select(cluster, gene, Top, p.value, FDR, everything()) %>%
  mutate(cluster = factor(cluster, levels = levels(colData(sce)$cluster)))

sce@metadata$marker_genes$scran$cluster <- marker_genes_cluster %>%
  filter(!is.na(Top), FDR <= 0.1)

glimpse(marker_genes_cluster)

marker_genes_cluster %>% head() %>% knitr::kable()
```

## Cell types

```{r marker_genes_cell_types}
marker_genes_cell_type <- findMarkers(sce, sce$cell_type_singler_blueprintencode_main)

for ( i in seq_along(marker_genes_cell_type) ) {
  marker_genes_cell_type[[i]] <- as.data.frame(marker_genes_cell_type[[i]]) %>%
    mutate(
      cell_type_singler_blueprintencode_main = names(marker_genes_cell_type)[i],
      gene = rownames(.)
    )
  marker_genes_cell_type[[i]][[paste0('logFC.', gsub(names(marker_genes_cell_type)[i], pattern = ' |\\-|\\+', replacement = '.'))]] <- 0
}

marker_genes_cell_type <- do.call(bind_rows, marker_genes_cell_type) %>%
  select(cell_type_singler_blueprintencode_main, gene, Top, p.value, FDR, everything()) %>%
  mutate(cell_type_singler_blueprintencode_main = factor(
      cell_type_singler_blueprintencode_main,
      levels = unique(cell_type_singler_blueprintencode_main)
    )
  )

sce@metadata$marker_genes$scran$cell_type_singler_blueprintencode_main <- marker_genes_cell_type %>%
  filter(!is.na(Top), FDR <= 0.1)

glimpse(marker_genes_cell_type)

marker_genes_cell_type %>% select(1:6) %>% head() %>% knitr::kable(booktabs = T)
```

# Export to Cerebro format

Finally, we use the `exportFromSeurat()` function of cerebroApp to export our Seurat object to a `.crb` file which can be loaded into Cerebro.

```{r export_data}
exportFromSCE(
  sce,
  assay = 'logcounts',
  file = paste0('cerebro_pbmc_SCE_', Sys.Date(), '.crb'),
  experiment_name = 'pbmc',
  organism = 'hg',
  columns_groups = c('sample','cluster','cell_type_singler_blueprintencode_main'),
  column_nUMI = 'sum',
  column_nGene = 'detected',
  add_all_meta_data = TRUE,
  verbose = FALSE
)
```

The Cerebro use interface can be launched using the `launchCerebro()` function.

# Session info

```{r session_info}
sessionInfo()
```
