---
title: "Benchmark of subsetting a `DelayedArray`"
author: "Roman Hillje"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Benchmark of subsetting a `DelayedArray`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Preparation

Load packages.

```{r message=FALSE}
library(tidyverse)
library(patchwork)
library(DelayedArray)
```

Load transcript counts in `DelayedArray` format.

```{r}
t <- readRDS('~/Dropbox/Cerebro_development/pbmc_Seurat_RleMatrix.crb')

expr <- t$expression
```

It contains `r format(nrow(expr), big.mark = ',')` genes (rows) and `r format(ncol(expr), big.mark = ',')` cells (columns).

To test whether either rows or columns are faster to access, we also prepare a transposed version of the array.

```{r}
expr_t <- Matrix::t(expr)
```

Define functions to subset array.

```{r}
get_col_means_direct <- function(matrix, gene_names) {
  return(Matrix::colMeans(matrix[ gene_names , ]))
}

get_col_means_direct_transposed <- function(matrix, gene_names) {
  return(Matrix::rowMeans(matrix[ , gene_names ]))
}

get_col_means_function <- function(matrix, gene_names) {
  indexes_of_genes_to_extract <- which(rownames(matrix) %in% gene_names)
  Matrix::colMeans(
    extract_array(
      matrix,
      list(
        indexes_of_genes_to_extract,
        NULL
      )
    )
  )
}

get_col_means_function_transposed <- function(matrix, gene_names) {
  indexes_of_genes_to_extract <- which(colnames(matrix) %in% gene_names)
  Matrix::rowMeans(
    extract_array(
      matrix,
      list(
        NULL,
        indexes_of_genes_to_extract
      )
    )
  )
}
```

# Perform benchmark

Measure the time it takes to calculate the mean expression across different numbers of randomly chosen genes (from 5 to 1000) using the four functions/methods defined above.
For each combination of the number of genes and function/method, 10 runs will be done.

```{r}
n_genes_to_test <- c(5,10,50,100,500,1000)
number_of_runs <- 10

results_extensive <- tibble(
  n_genes = numeric(),
  run = numeric(),
  method = character(),
  run_time = numeric()
)

for ( n_genes in n_genes_to_test ) {
  for ( run in seq(number_of_runs) ) {
    genes_to_extract <- rownames(expr)[sample(1:nrow(expr), size = n_genes)]
    results_extensive <- rbind(
      results_extensive,
      tribble(
        ~n_genes, ~run,               ~method,                                                                   ~run_time,
         n_genes,  run,              'direct',                system.time(get_col_means_direct(expr, genes_to_extract))[3],
         n_genes,  run,   'direct_transposed',   system.time(get_col_means_direct_transposed(expr_t, genes_to_extract))[3],
         n_genes,  run,            'function',              system.time(get_col_means_function(expr, genes_to_extract))[3],
         n_genes,  run, 'function_transposed', system.time(get_col_means_function_transposed(expr_t, genes_to_extract))[3]
      )
    )
  }
}
```

# Plot results

Plot access times.

```{r fig.height = 8, fig.width = 12, fig.align = "center"}
p1 <- results_extensive %>%
  group_by(n_genes, method) %>%
  summarise(
    mean = mean(run_time),
    sd = sd(run_time),
    se = sd/sqrt(number_of_runs),
    ci = qnorm(0.975)*sd/sqrt(number_of_runs)
  ) %>%
  ggplot(aes(n_genes, mean, group = method, color = method)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), color = 'black', width = .2, position = position_dodge(width = 0.25)) +
  geom_point(position = position_dodge(width = 0.25)) +
  scale_x_log10(name = 'Number of genes', breaks = n_genes_to_test, labels = scales::comma) +
  scale_y_continuous(name = 'Run time [seconds]') +
  annotation_logticks(sides = 'b') +
  theme_bw() +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank()
  )

p2 <- results_extensive %>%
  group_by(n_genes, method) %>%
  summarise(
    mean = mean(run_time),
    sd = sd(run_time),
    se = sd/sqrt(10),
    ci = qnorm(0.975)*sd/sqrt(10)
  ) %>%
  ggplot(aes(n_genes, mean, group = method, color = method)) +
  geom_smooth(method = lm, formula = y~x, se = FALSE) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), color = 'black', width = .2, position = position_dodge(width = 0.25)) +
  geom_point(position = position_dodge(width = 0.25)) +
  scale_x_log10(name = 'Number of genes', breaks = n_genes_to_test, labels = scales::comma) +
  scale_y_log10(name = 'Run time [seconds]') +
  annotation_logticks(sides = 'bl') +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank()
  )

p3 <- results_extensive %>%
  mutate(method_n_genes = paste0(method, '_', n_genes)) %>%
  ggplot(aes(n_genes, run_time, group = method_n_genes, color = method)) +
  geom_boxplot(position = position_dodge(width = 0.25)) +
  scale_x_log10(name = 'Number of genes', breaks = n_genes_to_test, labels = scales::comma) +
  scale_y_continuous(name = 'Run time [seconds]') +
  annotation_logticks(sides = 'b') +
  theme_bw() +
  theme(
    legend.position = 'none',
    panel.grid.minor = element_blank()
  )

p4 <- results_extensive %>%
  mutate(method_n_genes = paste0(method, '_', n_genes)) %>%
  ggplot(aes(n_genes, run_time, group = method_n_genes, color = method)) +
  geom_boxplot(position = position_dodge(width = 0.25)) +
  scale_x_log10(name = 'Number of genes', breaks = n_genes_to_test, labels = scales::comma) +
  scale_y_log10(name = 'Run time [seconds]') +
  annotation_logticks(sides = 'bl') +
  theme_bw() +
  theme(
    legend.position = 'none',
    panel.grid.minor = element_blank()
  )

p1 + p2 + p3 + p4 + plot_layout(ncol = 2)
```

# Add the stuff below

# Alternative array formats for expression data

## `HDF5Array`

Additional dependencies?

Create expression arrays in different formats.

```{r eval=FALSE}
#library('HDF5Array')

expression_dgCMatrix <- cerebro_seurat$expression@assays@data@listData$expression

expression_HDF5Array <- as(expression_dgCMatrix, "HDF5Array")
rownames(expression_HDF5Array) <- rownames(expression_dgCMatrix)
colnames(expression_HDF5Array) <- colnames(expression_dgCMatrix)

expression_RleArray <- as(expression_dgCMatrix, "RleArray")
rownames(expression_RleArray) <- rownames(expression_dgCMatrix)
colnames(expression_RleArray) <- colnames(expression_dgCMatrix)
```

Create `SCE` object with expression data in different formats.

```{r eval=FALSE}
test_sce_dgCMatrix <- SingleCellExperiment(list(counts = expression_dgCMatrix))
test_sce_HDF5Array <- SingleCellExperiment(list(counts = expression_HDF5Array))
test_sce_RleArray <- SingleCellExperiment(list(counts = expression_RleArray))
```

### Object size

```{r eval=FALSE}
print(object.size(expression_dgCMatrix), units = "auto", standard = "SI")
# 61.5 MB

print(object.size(expression_HDF5Array), units = "auto", standard = "SI")
# 1.5 MB

print(object.size(expression_RleArray), units = "auto", standard = "SI")
# 1.5 MB

print(object.size(test_sce_dgCMatrix), units = "auto", standard = "SI")
# 63.1 MB

print(object.size(test_sce_HDF5Array), units = "auto", standard = "SI")
# 3.2 MB

print(object.size(test_sce_RleArray), units = "auto", standard = "SI")
# 3.2 MB
```

The memory savings are impressive.

### Access speed

#### Single genes

```{r eval=FALSE}
genes <- 'A1BG'

system.time(as.vector(assay(test_sce_dgCMatrix[genes,], 'counts')))
#  user  system elapsed 
# 0.061   0.001   0.064

system.time(as.vector(counts(test_sce_dgCMatrix[genes,])))
#  user  system elapsed 
# 0.062   0.000   0.063

system.time(as.vector(assay(test_sce_HDF5Array[genes,], 'counts')))
#  user  system elapsed 
# 0.191   0.045   0.251

system.time(as.vector(counts(test_sce_HDF5Array[genes,])))
#  user  system elapsed 
# 0.184   0.043   0.232

system.time(as.vector(assay(test_sce_RleArray[genes,], 'counts')))
#  user  system elapsed 
# 0.067   0.027   0.104

system.time(as.vector(counts(test_sce_RleArray[genes,])))
#  user  system elapsed 
# 0.056   0.003   0.065
```

Quite a bit slower.

#### 1,000 genes

```{r eval=FALSE}
genes <- sample(rownames(test_sce_dgCMatrix), 1000)

system.time(as.vector(assay(test_sce_dgCMatrix[genes,], 'counts')))
#  user  system elapsed 
# 0.110   0.036   0.149

system.time(as.vector(counts(test_sce_dgCMatrix[genes,])))
#  user  system elapsed 
# 0.107   0.005   0.116

system.time(as.vector(assay(test_sce_HDF5Array[genes,], 'counts')))
#  user  system elapsed
# 1.790   0.481   2.334

system.time(as.vector(counts(test_sce_HDF5Array[genes,])))
#  user  system elapsed
# 1.733   0.438   2.197

system.time(as.vector(assay(test_sce_RleArray[genes,], 'counts')))
#  user  system elapsed
# 1.481   0.357   2.165

system.time(as.vector(counts(test_sce_RleArray[genes,])))
#  user  system elapsed
# 1.397   0.215   1.640
```

Also here there is a big difference.

### Store file

```{r eval=FALSE}
saveRDS(test_sce_dgCMatrix, 'test_sce_dgCMatrix.rds')
saveRDS(test_sce_HDF5Array, 'test_sce_HDF5Array.rds')
saveRDS(test_sce_RleArray, 'test_sce_RleArray.rds')

system("ls -alh | grep 'test_sce_'")
-rw-r--r--   1 romanhaa  staff    11M Aug 14 12:47 test_sce_dgCMatrix.rds
-rw-r--r--   1 romanhaa  staff   173K Aug 14 09:56 test_sce_HDF5Array.rds
-rw-r--r--   1 romanhaa  staff   8.9M Aug 14 09:56 test_sce_RleArray.rds
```

